---
title: "STOR 767 HW 3 - Computing"
author: "Brian N. White"
date: "10/27/2021"
output:
  html_document:
    df_print: paged
---

# Problem: IQ and successes

## Background: Measurement of Intelligence 

Case Study:  How intelligence relates to one's future successes?

**Data needed: `IQ.Full.csv`**

ASVAB (Armed Services Vocational Aptitude Battery) tests have been used as a screening test for those who want to join the army or other jobs. 

Our data set `IQ.csv` is a subset of individuals from the 1979 National Longitudinal Study of 
Youth (NLSY79) survey who were re-interviewed in 2006. Information about family, personal demographic such as gender, race and education level, plus a set of ASVAB (Armed Services Vocational Aptitude Battery) test scores are available. It is STILL used as a screening test for those who want to join the army! ASVAB scores were 1981 and income was 2005. 

**Our goals:** 

+ Is IQ related to one's successes measured by Income?
+ Is there evidence to show that Females are under-paid?
+ What are the best possible prediction models to predict future income? 


**The ASVAB has the following components:**

+ Science, Arith (Arithmetic reasoning), Word (Word knowledge), Parag (Paragraph comprehension), Numer (Numerical operation), Coding (Coding speed), Auto (Automative and Shop information), Math (Math knowledge), Mechanic (Mechanic Comprehension) and Elec (Electronic information).
+ AFQT (Armed Forces Qualifying Test) is a combination of Word, Parag, Math and Arith.
+ Note: Service Branch requirement: Army 31, Navy 35, Marines 31, Air Force 36, and Coast Guard 45,(out of 100 which is the max!) 

**The detailed variable definitions:**

Personal Demographic Variables: 

 * Race: 1 = Hispanic, 2 = Black, 3 = Not Hispanic or Black
 * Gender: a factor with levels "female" and "male"
 * Educ: years of education completed by 2006
 
Household Environment: 
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent?™s household regularly read
	magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent?™s household regularly read
	newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent?™s household had a library card
	in 1979, otherwise 0
* MotherEd: mother's years of education
* FatherEd: father's years of education

Variables Related to ASVAB test Scores in 1981 (Proxy of IQ's)

* AFQT: percentile score on the AFQT intelligence test in 1981 
* Coding: score on the Coding Speed test in 1981
* Auto: score on the Automotive and Shop test in 1981
* Mechanic: score on the Mechanic test in 1981
* Elec: score on the Electronics Information test in 1981

* Science: score on the General Science test in 1981
* Math: score on the Math test in 1981
* Arith: score on the Arithmetic Reasoning test in 1981
* Word: score on the Word Knowledge Test in 1981
* Parag: score on the Paragraph Comprehension test in 1981
* Numer: score on the Numerical Operations test in 1981

Variable Related to Life Success in 2006

* Income2005: total annual income from wages and salary in 2005. We will use a natural log transformation over the income.

The following 10 questions are answered as 1: strongly agree, 2: agree, 3: disagree, 4: strongly disagree

* Esteem 1: I am a person of worth?
* Esteem 2: I have a number of good qualities?
* Esteem 3: I am inclined to feel like a failure?
* Esteem 4: I do things as well as others?
* Esteem 5: I do not have much to be proud of?
* Esteem 6: I take a positive attitude towards myself and others?
* Esteem 7: I am satisfied with myself?
* Esteem 8: I wish I could have more respect for myself?
* Esteem 9: I feel useless at times?
* Esteem 10: I think I am no good at all?

**Note: we will not use the Esteem scores in this homework.**

```{r, include=FALSE}
# Loading and cleaning of data.
iq <- read.csv("IQ.Full.csv")
iq[nrow(iq), ]
names(iq)
iq$Imagazine <- as.factor(iq$Imagazine)
iq$Inewspaper <- as.factor(iq$Inewspaper) 
iq$Ilibrary <- as.factor(iq$Ilibrary)
iq$Race <- as.factor(iq$Race)
iq$Gender <- as.factor(iq$Gender)
iq$Income2005 <- log(iq$Income2005)
iq <- iq[, -1]
iq <- iq[, -c(22:31)]
str(iq)
index1 <- sample(nrow(iq), 200)
iq1 <- iq[index1, ]
str(iq1)
summary(iq1)
cor(iq[, 11:21]) # ASVAB tests are highly correlated
pairs(iq[, 11:21])
```

```{r load packages, include=FALSE}
pacman::p_load(randomForest, tree, ISLR, rpart, rattle, pROC, partykit, ggplot2, glmnet, lda, class, tidyverse)
```

## 1. EDA
i. Make a few informative tables or plots to see how intelligence might relate to the log income.
ii. Make a training data and testing data (approximately 2/3 observations as training data)

**Solution**

**i**
```{r EDA part i}
iq %>%
  ggplot(aes(AFQT, Income2005, col=Educ)) +
  geom_point() +
  facet_wrap(~Race)

iq %>%
  group_by(Race) %>%
  summarize(n=n(), proportion=n()/2584)
```
**ii**

```{r EDA part ii}
set.seed(5)
#generate training data indices 
iq_train_index <- sample(nrow(iq), (2/3)*nrow(iq))
#generate training data with 2/3 of original observations
iq_train <- iq[iq_train_index,]
#generate testing data from the left over observations
iq_test <- iq[-iq_train_index,]
```


## 2. Single trees
i. Build a reasonable single tree using all the variables to predict log income. Display the tree.
ii. Explain the predictions. Does that agree with our intuition?
iii. Can you detect gender bias against women from the tree built?
iv. Do we have evidence that the high IQs result in higher income? What is the testing error for the tree you built?

**Solution**

**i**

```{r}
tree_train <- tree(Income2005~., iq_train)
plot(tree_train)
text(tree_train, pretty = 1)
```
**ii**

The decision tree tells us that men with more than 15.5 years of education are the highest earning demographic overall with an expected salary of ~70,000 USD. In contrast, men with no more than 15.5 years of education who score no more than 15.5 on the arithmetic exam and whose parents had a combined income of no more than 10,000 USD are the lowest earning demographic among men, with an expected salary of ~21120 USD. 

Similarly, women with more than 15.5 years of education are the highest earning demographic among women, with an expected salary of ~36690 USD. In contrast, women with no more than 15.5 years of education who performed no better than the ~18th percentile of AFQT test takers are the lowest earning demographic overall, with an expected salary of ~13240 USD. 

Similar analysis can be performed for the remaining paths on the decision tree via inspection of the tree plot above. 

These results are consistent with intuition in the sense that an observation's intelligence, to that extent one believes the variables in the tree are reasonable measures of intelligence, predicts a higher salary. 

**iii**

The decision tree appears to indicate the presence of gender bias. For instance, women with more than 15.5 years of education have an expected salary of ~36690 USD whereas men with more than 15.5 years of education have an expected salary of ~70000 USD. 

**iv**

As discussed in part i, the decision tree offers evidence for the claim that high IQs result in higher income. The testing error (MSE) for the tree in question is ~0.791

```{r}
tree_test <- predict(tree_train, iq_test)

#testing error (MSE), note denominator is n - (# of terminal nodes)
tree_test_err <- sum((iq_test$Income2005-tree_test)^2)/855
tree_test_err
```

## 3. Bagging with two bootstrap trees.
i. Build and display two bootstrap trees. Commenting on the differences among the two trees.
ii. Bag the twos tree by taking average of the two bootstrap trees above. What is the testing error?

**Solution**

**i**

The bootstrap trees are generated and displayed via the code-chunk below. The two trees differ in a number of ways. First, tree_boot1 used 8 of the available variables whereas tree_boot2 used 6. Further, tree_boot1 has 10 terminal nodes whereas tree_boot2 has 9. They fit the training data comparably well (as would be expected) with residual mean deviance of .6358 and .6904, respectively. The tree-diagram for tree_boot1 indicates that the following observations are predicted to have the greatest salaries: men who performed above the 20th percentile on the AFQT, have over 15 years of education and grew up in a family with an income greater than 2450 have predicted salaries of about 40K. In contrast, women with arithmetic test scores greater than 15.5 are predicted to have salaries of about 55K. A similar analysis can be readily performed through inspection of the tree-diagram for tree_boot2.

```{r generate two trees via bootstrap sampling}
set.seed(5)
#generate two bootstrap samples 
sample.df <- function(df){
  index <- sample(nrow(df), nrow(df), replace=T)
  sample <- df[index,]
  return(sample)
}

boot1 <- sample.df(iq_train)
boot2 <- sample.df(iq_train)

#generate two bootstrap trees
tree_boot1<- tree(Income2005~., boot1)
tree_boot2<- tree(Income2005~., boot2)

#summary of trees
summary(tree_boot1)
summary(tree_boot2)

#display the two bootstrap trees
plot(tree_boot1)
text(tree_boot1, pretty = 1)
plot(tree_boot2)
text(tree_boot2, pretty = 1)
```

**ii**

```{r bag the trees}
#compute the test predictions for each tree
tree_boot1_test <- predict(tree_boot1, iq_test)
tree_boot2_test <- predict(tree_boot2, iq_test)

#bag the trees by taking the average of their respective test data predictions
bagged_predictions <- (tree_boot1_test + tree_boot2_test)/2

#compute testing error (MSE) for bagged tree predictions
bagged_test_err <- sum((iq_test$Income2005-bagged_predictions)^2)/852
bagged_test_err
```

## 4. Build the best possible random forest
i. Show the process how you tune the number of the trees and `mtry`
ii. What is the testing error for your random forest?
iii. Comment on the pro and cons for the random forest comparing with the single tree you built earlier.

**Solution**

**i**

```{r}

```

